Evaluation of 3-point correlation functions from structural images on CPU and GPU architectures: accounting for anisotropy effects

### 2024-02-12 paper.tex
An invaluable contribution from Alexey Khlupin

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts.  Moreover, 3-point and CFs with $n>3$
are believed to possess larger information content and provide more information
about studies structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of sample are orders of
magnitude lower than existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun(accepted)]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

    functions \cite{Degeneraty.045306,cherkasov2023towards};

@@

\section{Methods}
\label{sec:methods}
\subsection{3-point correlation functions}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^n$, where the index can be interpreted as a phase on the
image $A$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

Now we can define the three-point correlation function
$S_3: \mathbb{R}^{2n} \rightarrow [0, 1]$:
\begin{equation}
  S_3^n(x_1, x_2) = \langle I_n(x) I_n(x + x_1) I_n(x + x_2) \rangle
\end{equation}
where $\langle \dots \rangle$ means average over all points in $A$. It can be
understood as the probability that all three vertices of a triangle
$(0, x_1, x_2)$ will appear in the subset $A_n$ when the triangle is randomly
thrown in $A$. When $A$ represent isotropic medium, $x_1$ and $x_2$ can be
replaced with three numbers $r_1 = |x_1|$, $r_2 = |x_2|$ and $r_3 = |x_2 - x_1|$
to get a function of three scalar arguments.

Another function closely related to the three-point correlation function is the
cluster function which can be defined as:
\begin{equation}
  C_3^n(x_1, x_2) = \sum_{k=1}^K \langle \tilde{I}_k(x) \tilde{I}_k(x + x_1)
  \tilde{I}_k(x + x_2) \rangle
\end{equation}
where $\tilde{I}_k$ is an indicator function for $k$-th cluster and $K$ is a
total number of clusters. A cluster is a subset of $A_n$ so that any two points
in this subset can be connected with a curve belonging to that subset.

By introducing a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function, we can define three additional
correlation functions -- these are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as:
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_v(x + x_1) I_v(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_v(x + x_1) I_v(x + x_2) \rangle
\end{align}
Here $I_v$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g., can be measured in $\mu m^{-3}$, $\mu m^{-2}$ and
$\mu m^{-1}$ or other convenient dimensionless units depending on the problem to be solved). 
These functions describe the interface of the subset $A_n$ and
its spatial configuration with respect to $A_v$.

@@

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator:
\begin{equation}
  A[x] \leftarrow A[x+z \mod d]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in Appendix
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.4\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.4\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

@@

\begin{figure*}[tp]
  \centering
  \subfigure[$C_3$]{
    \includegraphics[width=0.4\linewidth]{images/c3-map.png}}
  \hfill
  \subfigure[$F_{sss}$]{
    \includegraphics[width=0.4\linewidth]{images/surf3-map.png}}
  \vskip\baselineskip
  \subfigure[$F_{ssv}$]{
    \includegraphics[width=0.4\linewidth]{images/surf2void-map.png}}
  \hfill
  \subfigure[$F_{svv}$]{
    \includegraphics[width=0.4\linewidth]{images/surfvoid2-map.png}}
  \caption[]{Correlation maps for the sample of soil.}
  \label{fig:soil-maps}
\end{figure*}

@@

We mentioned that classic set of correlation functions describes both the
geometry and topology of the structure at hand -- but this is to yet unknown
extent. With the advent of analytical methods \cite{cherkasov2023towards} to
assess information content of arbitrary descriptor for structure of any size, it
will be possible to choose the most useful set of correlation functions and
compare their information content against classical and other metrics. If
computationally effective methodology for information content assessment will be
developed, it should amend all aforementioned uncertainties and provide a clear
winner -- a plethora of different 2-point CFs or a less numerous set of
higher-order functions. Such long awaited answer would not be universal thought
and depends not only on descriptor's information content versus data volume
ratio, but also on its applicability to solve particular problems. For example,
it is well known that cluster $C_2$ function provides a lot of information in
addition to $S_2$ \cite{JiaoPNAS}, yet it's application in stochastic
reconstructions is very limited due to computational burden of cluster
relabeling during annealing \cite{PLoS_ONE}. Still, effective information
content assessment techniques will provide a possibility to account for such
additional application constraints.

>>>>

\begin{abstract}
Structures, or spatial arrangements of matter and energy, including some fields
(e.g., velocity or pressure) are ubiquitous in research applications and
frequently require description for subsequent analysis, or stochastic
reconstruction from limited data.  The classical descriptors are 2-point
correlation functions (CFs), but the computation of 3-point statistics is known
to be advantageous in some cases as they can probe non-Gaussian signatures, not
captured by their 2-point counterparts. Moreover, $n$-point CFs with $n \ge 3$
are believed to possess larger information content and provide more information
about studies structures. In this paper, we have developed algorithms and code
to compute $S_3, C_3, F_{sss}, F_{ssv}$ and $F_{svv}$ with a right-angle
triangle pattern. This pattern is much easier to vectorize and parallelize
compared to previously used random triangle or spider-web sampling
patterns. Moreover, it allows to easily implement computations of directional
3-point CFs. The execution times for the same number of sample are orders of
magnitude lower than existing published counterparts. We showed that the volume
of data produced get unwieldy very easily, especially if computations are
performed in frequency domain. For these reasons until information content of
different sets of correlation functions with different "n-pointness" is not
known, advantages of CFs with $n>3$ are not clear. Nonetheless, developed
algorithms and code are universal enough to be easily extendable to any $n$
(with increasing computational and RAM burden). All results are available as
part of open source package \code{CorrelationFunctions.jl} [Postnicov et al.,
  2024, Comp.Phys.Commun(accepted)]. As described in this paper, 3-point CFs
computations can be immediately applied in a great number of research
applications, for example: 1) flow and transport velocity fields analysis or any
data with non-Gaussian signatures, 2) deep learning for structural and physical
properties, 3) structure taxonomy and categorization.  In all these and numerous
other potential cases the ability to compute directional 3-point functions may
be crucial. Notably, the organization of the code functions allows computation
of cross-correlation, i.e., one can compute 3-point CFs for multiphase images
(while binary structures were used in this paper for simplicity of
explanations).
\end{abstract}

@@

    functions \cite{Degeneraty.045306,CHERKASOV2024129400};

@@

\section{Methods}
\label{sec:methods}
\subsection{3-point correlation functions}
Let $A_1, A_2, \dots, A_n$ be pairwise disjoint subsets of a set
$A \subset \mathbb{R}^m$, where the index can be interpreted as a phase on the
image $A$. We introduce a family of indicator functions
$I_n(x) : A \rightarrow \{0,1\}$ defined as:
\begin{equation}
  I_n(x) = \left\{
  \begin{array}{ll}
    1 & \quad x \in A_n \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}

Now we can define the three-point correlation function
$S_3: \mathbb{R}^{2m} \rightarrow [0, 1]$:
\begin{equation}
  S_3^n(x_1, x_2) = \langle I_n(x) I_n(x + x_1) I_n(x + x_2) \rangle
\end{equation}
where $\langle \dots \rangle$ means ensemble average. It can be understood as
the probability that all three vertices of a triangle $(0, x_1, x_2)$ will
appear in the subset $A_n$ when the triangle is randomly thrown in $A$. When $A$
represent isotropic medium, $x_1$ and $x_2$ can be replaced with three numbers
$r_1 = |x_1|$, $r_2 = |x_2|$ and $r_3 = |x_2 - x_1|$ to get a function of three
scalar arguments.

Another function closely related to the three-point correlation function is the
cluster function which can be defined as:
\begin{equation}
  C_3^n(x_1, x_2) = \sum_{k=1}^K \langle I_n^k(x) I_n^k(x + x_1) I_n^k(x + x_2) \rangle
\end{equation}
where $I_n^k$ is an indicator function for $k$-th cluster of $A_n$ and $K$
is a total number of clusters. A cluster is a subset of $A_n$ so that any two
points in this subset can be connected with a curve belonging to that subset.

By introducing a function $M_n(x) = |\nabla I_n(x)|$ which is a so-called interface
indicator function, we can define three additional
correlation functions -- these are the surface-surface-surface function
($F_{sss}$), surface-surface-void function ($F_{ssv}$) and surface-void-void
function ($F_{svv}$). They are defined as:
\begin{align}
  F_{sss}^n(x_1, x_2) &= \langle M_n(x) M_n(x + x_1) M_n(x + x_2) \rangle \\
  F_{ssv}^n(x_1, x_2) &= \langle M_n(x) M_{(void)}(x + x_1) I_{(void)}(x + x_2) \rangle \\
  F_{svv}^n(x_1, x_2) &= \langle M_n(x) I_{(void)}(x + x_1) I_{(void)}(x + x_2) \rangle
\end{align}
Here $I_{(void)}$ is an indicator function for the subset of the void phase
$A_v$. Unlike $S_3$ and $C_3$ functions, these functions do not have the meaning
of probability and have the units of measurement inversely proportional to
volume, surface and length, respectively (e.g., can be measured in $\mu m^{-3}$,
$\mu m^{-2}$ and $\mu m^{-1}$ or other convenient dimensionless units depending
on the problem to be solved).  These functions describe the interface of the
subset $A_n$ and its spatial configuration with respect to $A_v$.

@@

There are two common boundary conditions when applying $\mathfrak{S}$. The
first condition is to extend $A_n$ periodically (\cref{fig:s3-periodic}) when
accessing out-of-bounds array elements. In this case $\mathfrak{S}_z$ is a
circular shift operator (here ``$x \mod y$'' means a vector
$(x_1 \mod y_1, x_2 \mod y_2, \dots)$):
\begin{equation}
  A[x] \leftarrow A[x+z \mod shape(A)]
\end{equation}
In this case $Norm(A, x_1, x_2)$ does not depend on $x_1$ and $x_2$ and
is equal to the total number of elements in $A$. The second condition is to
replace out-of-bound elements with zeros (so-called zero padding)
(\cref{fig:s3-zeros}). A formula for $Norm(A, x_1, x_2)$ in this case is in Appendix
\ref{sec:number-of-trials} and $\mathfrak{S}_z$ is defined as follows:
\begin{equation}
  A[x] \leftarrow \left\{
  \begin{array}{ll}
    A[x+z] & \quad \text{if $x+z$ is a correct index in $A$} \\
    0 & \quad \text{otherwise}
  \end{array}
  \right.
\end{equation}
\begin{figure*}[tp]
  \centering
  \subfigure[Periodic boundary conditions]{
    \includegraphics[width=0.4\linewidth]{images/periodic.png}
    \label{fig:s3-periodic}}
  \hfill
  \subfigure[Zero padding]{
    \includegraphics[width=0.4\linewidth]{images/zeros.png}
    \label{fig:s3-zeros}}
  \caption[]{Computation of unnormalized $S_3$ function at point $(1, 2)$ for
    one-dimensional sequence of length 8.}
  \label{fig:s3-computation}
\end{figure*}

@@

\begin{figure*}[tp]
  \centering
  \subfigure[$C_3$]{
    \includegraphics[width=0.4\linewidth]{images/c3-map.png}}
  \hfill
  \subfigure[$F_{sss}$]{
    \includegraphics[width=0.4\linewidth]{images/surf3-map.png}}
  \vskip\baselineskip
  \subfigure[$F_{ssv}$]{
    \includegraphics[width=0.4\linewidth]{images/surf2void-map.png}}
  \hfill
  \subfigure[$F_{svv}$]{
    \includegraphics[width=0.4\linewidth]{images/surfvoid2-map.png}}
  \caption[]{Correlation maps for the cluster function and the surface functions
    of the sample of soil. The functions are calculated for a rotation of the
    sample which aligns the direction of anisotropy with the axes.}
  \label{fig:soil-maps}
\end{figure*}

@@

We mentioned that classic set of correlation functions describes both the
geometry and topology of the structure at hand -- but this is to yet unknown
extent. With the advent of analytical methods \cite{CHERKASOV2024129400} to
assess information content of arbitrary descriptor for structure of any size, it
will be possible to choose the most useful set of correlation functions and
compare their information content against classical and other metrics. If
computationally effective methodology for information content assessment will be
developed, it should amend all aforementioned uncertainties and provide a clear
winner -- a plethora of different 2-point CFs or a less numerous set of
higher-order functions. Such long awaited answer would not be universal thought
and depends not only on descriptor's information content versus data volume
ratio, but also on its applicability to solve particular problems. For example,
it is well known that cluster $C_2$ function provides a lot of information in
addition to $S_2$ \cite{JiaoPNAS}, yet it's application in stochastic
reconstructions is very limited due to computational burden of cluster
relabeling during annealing \cite{PLoS_ONE}. Still, effective information
content assessment techniques will provide a possibility to account for such
additional application constraints.
